


Question: How have the restarurant industries grown (or shrunk?) over the past 10 years in Seattle, San Francisco, and DC?

to do:
- combine total employment for all workers (bartenderes, servers, cooks, dishwashers) for each year
- make a column for each city and combine into one dataframe
- graph dataframe to compare employment across cities

1. Read in all data frames
```{r}
sea_bartenders <- read_csv("data/sea_bartenders.csv")
sea_cooks <- read_csv("data/sea_cooks.csv")
sea_dishwashers <- read_csv("data/sea_dishwashers.csv")
sea_servers <- read_csv("data/sea_servers.csv")


dc_bartenders <- read_csv("data/dc_bartenders.csv")
dc_cooks <- read_csv("data/dc_cooks.csv")
dc_dishwashers <- read_csv("data/dc_dishwashers.csv")
dc_servers <- read_csv("data/dc_servers.csv")


sf_bartenders <- read_csv("data/sf_bartenders.csv")
sf_cooks <- read_csv("data/sf_cooks.csv")
sf_dishwashers <- read_csv("data/sf_dishwashers.csv")
sf_servers <- read_csv("data/sf_servers.csv")

```





1. Create dataframes with each total employment for DC workers and combine
```{r}

#DC 
dc_bartenders_emp <- dc_bartenders %>%
  select(year, tot_emp) %>%
  rename(dc_bartenders_emp = tot_emp)

dc_servers_emp <- dc_servers %>%
  select(year, tot_emp) %>%
  rename(dc_servers_emp = tot_emp)

dc_dishwashers_emp <- dc_dishwashers %>%
  select(year, tot_emp) %>%
  rename(dc_dishwashers_emp = tot_emp)

dc_cooks_emp <- dc_cooks %>%
  select(year, tot_emp) %>%
  rename(dc_cooks_emp = tot_emp)

dc_tot_emp <- dc_bartenders_emp %>%
  left_join(dc_servers_emp) %>%
  left_join(dc_dishwashers_emp) %>%
  left_join(dc_cooks_emp)


#Seattle 
sea_bartenders_emp <- sea_bartenders %>%
  select(year, tot_emp) %>%
  rename(sea_bartenders_emp = tot_emp)

sea_servers_emp <- sea_servers %>%
  select(year, tot_emp) %>%
  rename(sea_servers_emp = tot_emp)

sea_dishwashers_emp <- sea_dishwashers %>%
  select(year, tot_emp) %>%
  rename(sea_dishwashers_emp = tot_emp)

sea_cooks_emp <- sea_cooks %>%
  select(year, tot_emp) %>%
  rename(sea_cooks_emp = tot_emp)

sea_tot_emp <- sea_bartenders_emp %>%
  left_join(sea_servers_emp) %>%
  left_join(sea_dishwashers_emp) %>%
  left_join(sea_cooks_emp)



#SanFran
sf_bartenders_emp <- sf_bartenders %>%
  select(year, tot_emp) %>%
  rename(sf_bartenders_emp = tot_emp)

sf_servers_emp <- sf_servers %>%
  select(year, tot_emp) %>%
  rename(sf_servers_emp = tot_emp)

sf_dishwashers_emp <- sf_dishwashers %>%
  select(year, tot_emp) %>%
  rename(sf_dishwashers_emp = tot_emp)

sf_cooks_emp <- sf_cooks %>%
  select(year, tot_emp) %>%
  rename(sf_cooks_emp = tot_emp)

sf_tot_emp <- sf_bartenders_emp %>%
  left_join(sf_servers_emp) %>%
  left_join(sf_dishwashers_emp) %>%
  left_join(sf_cooks_emp)
```


2. Create a new column that adds all together and limit dataframe to those columns to combine
```{r}
dc_tot_emp <- dc_tot_emp %>%
  mutate(dc_tot_emp = dc_bartenders_emp + dc_servers_emp + dc_dishwashers_emp + dc_cooks_emp ) %>%
  select(year, dc_tot_emp)

sea_tot_emp <- sea_tot_emp %>%
  mutate(sea_tot_emp = sea_bartenders_emp + sea_servers_emp + sea_dishwashers_emp + sea_cooks_emp ) %>%
  select(year, sea_tot_emp)

sf_tot_emp <- sf_tot_emp %>%
  mutate(sf_tot_emp = sf_bartenders_emp + sf_servers_emp + sf_dishwashers_emp + sf_cooks_emp ) %>%
  select(year, sf_tot_emp)
```


3. Merge all dataframes together again :)
```{r}
tot_emp <- dc_tot_emp %>%
  left_join(sea_tot_emp) %>%
  left_join(sf_tot_emp)
```


4. Write to a csv to make a graph in DataWrapper
```{r}
write.csv(tot_emp, "tot_emp.csv")
```


Final graph: https://www.datawrapper.de/_/uInne/ 

