Question #5: How have the restaurant industries grown (or shrunk?) over the past 10 years in Seattle, San Francisco, and DC?

Answer: To answer this question, I graphed the total employment numbers for each cities over the past 10 years:


Graph of total employment numbers: https://www.datawrapper.de/_/uInne/ 
Graph of total employment (rate per 1000 people): https://www.datawrapper.de/_/KyK5y/ 

As seen in these charts, the restaurant industry in San Francisco had been steady (but at a slow decline). The restaurant industry in DC was increasing slowly and the restaurant industry in Seattle continued at a steady rate. However, the restaurant industry employment in all three cities tanked when COVID hit in 2020. Knowing this can make it easier to understand potential wage increases and decreases, especially during COVID when employment shrunk in all three cities. 



1. Create dataframes with each total employment for DC workers and combine
```{r}

#DC 
dc_bartenders_emp <- dc_bartenders %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(dc_bartenders_emp = tot_emp) %>%
  rename(dc_bartenders_emp_per_1000 = jobs_1000)

dc_servers_emp <- dc_servers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(dc_servers_emp = tot_emp) %>%
  rename(dc_servers_emp_per_1000 = jobs_1000)

dc_dishwashers_emp <- dc_dishwashers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(dc_dishwashers_emp = tot_emp) %>%
  rename(dc_dishwashers_emp_per_1000 = jobs_1000)

dc_cooks_emp <- dc_cooks %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(dc_cooks_emp = tot_emp) %>%
  rename(dc_cooks_emp_per_1000 = jobs_1000)

dc_tot_emp <- dc_bartenders_emp %>%
  left_join(dc_servers_emp) %>%
  left_join(dc_dishwashers_emp) %>%
  left_join(dc_cooks_emp)


#Seattle 
sea_bartenders_emp <- sea_bartenders %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sea_bartenders_emp = tot_emp) %>%
  rename(sea_bartenders_emp_per_1000 = jobs_1000)

sea_servers_emp <- sea_servers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sea_servers_emp = tot_emp) %>%
  rename(sea_servers_emp_per_1000 = jobs_1000)

sea_dishwashers_emp <- sea_dishwashers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sea_dishwashers_emp = tot_emp) %>%
  rename(sea_dishwashers_emp_per_1000 = jobs_1000)

sea_cooks_emp <- sea_cooks %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sea_cooks_emp = tot_emp) %>%
  rename(sea_cooks_emp_per_1000 = jobs_1000)

sea_tot_emp <- sea_bartenders_emp %>%
  left_join(sea_servers_emp) %>%
  left_join(sea_dishwashers_emp) %>%
  left_join(sea_cooks_emp)



#SanFran
sf_bartenders_emp <- sf_bartenders %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sf_bartenders_emp = tot_emp) %>%
  rename(sf_bartenders_emp_per_1000 = jobs_1000)

sf_servers_emp <- sf_servers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sf_servers_emp = tot_emp) %>%
  rename(sf_servers_emp_per_1000 = jobs_1000)

sf_dishwashers_emp <- sf_dishwashers %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sf_dishwashers_emp = tot_emp) %>%
  rename(sf_dishwashers_emp_per_1000 = jobs_1000)

sf_cooks_emp <- sf_cooks %>%
  select(year, tot_emp, jobs_1000) %>%
  rename(sf_cooks_emp = tot_emp) %>%
  rename(sf_cooks_emp_per_1000 = jobs_1000)

sf_tot_emp <- sf_bartenders_emp %>%
  left_join(sf_servers_emp) %>%
  left_join(sf_dishwashers_emp) %>%
  left_join(sf_cooks_emp)
```


2. Create a new column that adds all together and limit dataframe to those columns to combine
```{r}
dc_tot_emp <- dc_tot_emp %>%
  mutate(dc_tot_emp = dc_bartenders_emp + dc_servers_emp + dc_dishwashers_emp + dc_cooks_emp ) %>%
  mutate(dc_tot_emp_per_1000 = dc_bartenders_emp_per_1000 + dc_servers_emp_per_1000 + dc_dishwashers_emp_per_1000 + dc_cooks_emp_per_1000 ) %>%
  select(year, dc_tot_emp, dc_tot_emp_per_1000)

sea_tot_emp <- sea_tot_emp %>%
  mutate(sea_tot_emp = sea_bartenders_emp + sea_servers_emp + sea_dishwashers_emp + sea_cooks_emp ) %>%
  mutate(sea_tot_emp_per_1000 = sea_bartenders_emp_per_1000 + sea_servers_emp_per_1000 + sea_dishwashers_emp_per_1000 + sea_cooks_emp_per_1000 ) %>%
  select(year, sea_tot_emp, sea_tot_emp_per_1000)

sf_tot_emp <- sf_tot_emp %>%
  mutate(sf_tot_emp = sf_bartenders_emp + sf_servers_emp + sf_dishwashers_emp + sf_cooks_emp ) %>%
  mutate(sf_tot_emp_per_1000 = sf_bartenders_emp_per_1000 + sf_servers_emp_per_1000 + sf_dishwashers_emp_per_1000 + sf_cooks_emp_per_1000 ) %>%
  select(year, sf_tot_emp, sf_tot_emp_per_1000)
```


3. Merge all dataframes together again :)
```{r}
tot_emp <- dc_tot_emp %>%
  left_join(sea_tot_emp) %>%
  left_join(sf_tot_emp)
```


4. Write to a csv to make a graph in DataWrapper
```{r}
write.csv(tot_emp, "tot_emp.csv")
```

